---
title: "Fitting Marginal Cox Models with Clustered Interval-Censored Data"
author: ""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting Marginal Cox Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(BCPP.Package)
```

## Overview

This vignette demonstrates how to fit marginal Cox models to clustered, interval-censored data using the composite_coxIC() function. We highlight an example using time-dependent covariates with a pre-generated dataset containing 200 clusters and four strata.
This package also show how to generate simulated data that mimics the Bostwana Combination Prevention Project(BCPP) structure, using included data generation utilities.

## Getting started with `{intervalCRT}`
```{r}
library(BCPP.Package)
```


## Simulating Clustered Interval-Censored Data

The following functions allow you to simulate baseline hazards and clustered, interval-censored survival data. For example, we fist simulate a dataset with

- 200 clusters of 20–30 individuals

- 4 strata

- 2 covariates: treat and utt

```{r}
set.seed(112358)
baseline <- lapply(1:4, function(i) return_baseline(320, 8))

data <- gen_time_varying_cov(
  n.groups = 200,
  group.size.range = c(20, 30),
  beta = c(-1, 0.5, 0),
  visits.generator = gen_visits,
  censoring.rate = 0,
  rho = 0.5,
  baseline = baseline,
  S = function(s) sample(1:4, s, replace = TRUE)
)

head(data)
```

